---
title: "Clean Process 2010 - 2023 ACS IPUMS Data"
output: html_document
date: "2025-04-02"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries and data
```{r}
library(tidyverse)
library(readxl)
library(readr)
library(data.table)
library(dplyr)
library(janitor)

data <- fread("G:/My Drive/INFO 658 - Information Visualization/Tableau Report/Raw 2010 - 2023 ACS IPUMS Data.csv")
## "G:/My Drive/INFO 658 - Information Visualization/Tableau Report/Raw 2010 - 2023 ACS IPUMS Data.csv"
## "~/Desktop/Raw 2010 - 2023 ACS IPUMS Data.csv"
```

# Take a quick look at the data
```{r}
head(data, 10)
glimpse(data)
```

# Remove any empty observations
```{r}
data2 <- data %>% drop_na()
glimpse(data2)
```

# Rename region column and variables
NORTHEAST REGION
- 11: New England Division
- 12: Middle Atlantic Division
- 13: Mixed Northeast Divisions (1970 Metro)
MIDWEST REGION
- 21: East North Central Div.
- 22: West North Central Div.
- 23: Mixed Midwest Divisions (1970 Metro)
SOUTH REGION
- 31: South Atlantic Division
- 32: East South Central Div.
- 33: West South Central Div.
- 34: Mixed Southern Divisions (1970 Metro)
WEST REGION
- 41: Mountain Division
- 42: Pacific Division
- 43: Mixed Western Divisions (1970 Metro)
STATE UNKNOWN
- 91: Military/Military reservations
- 92: PUMA boundaries cross state lines-1% sample
- 97: State not identified
- 99: Not identified
```{r}
unique(data2$REGION)
data2 <- data2 %>%
  mutate(
    Region = case_when(
      REGION %in% c(11, 12, 13) ~ "Northeast",
      REGION %in% c(21, 22, 23) ~ "Midwest",
      REGION %in% c(31, 32, 33, 34) ~ "South",
      REGION %in% c(41, 42, 43) ~ "West",
      TRUE ~ "State Unknown"
    )
  )
table(data2$Region)
```

# Rename state column and variables
```{r}
unique(data2$STATEFIP)
statename <- c(
  "1" = "Alabama", "2" = "Alaska", "4" = "Arizona", "5" = "Arkansas", "6" = "California",
  "8" = "Colorado", "9" = "Connecticut", "10" = "Delaware", "11" = "District of Columbia",
  "12" = "Florida", "13" = "Georgia", "15" = "Hawaii", "16" = "Idaho", "17" = "Illinois",
  "18" = "Indiana", "19" = "Iowa", "20" = "Kansas", "21" = "Kentucky", "22" = "Louisiana",
  "23" = "Maine", "24" = "Maryland", "25" = "Massachusetts", "26" = "Michigan", "27" = "Minnesota",
  "28" = "Mississippi", "29" = "Missouri", "30" = "Montana", "31" = "Nebraska", "32" = "Nevada",
  "33" = "New Hampshire", "34" = "New Jersey", "35" = "New Mexico", "36" = "New York",
  "37" = "North Carolina", "38" = "North Dakota", "39" = "Ohio", "40" = "Oklahoma",
  "41" = "Oregon", "42" = "Pennsylvania", "44" = "Rhode Island", "45" = "South Carolina",
  "46" = "South Dakota", "47" = "Tennessee", "48" = "Texas", "49" = "Utah", "50" = "Vermont",
  "51" = "Virginia", "53" = "Washington", "54" = "West Virginia", "55" = "Wisconsin", "56" = "Wyoming"
)
data2$State <- statename[as.character(data2$STATEFIP)]
table(data2$State)
```

# Rename sex and race columns and variables
```{r}
unique(data2$SEX)
data2 <- data2 %>%
  mutate(
    Sex = case_when(
      SEX == 1 ~ "Male",
      SEX == 2 ~ "Female",
      TRUE ~ "Missing/Blank"
    )
  )
table(data2$Sex)

unique(data2$RACE)
data2 <- data2 %>%
  mutate(
    Race = case_when(
      RACE == 1 ~ "White",
      RACE == 2 ~ "Black/African American",
      RACE == 3 ~ "American Indian or Alaska Native",
      RACE %in% c(4, 5, 6) ~ "Asian or Pacific Islander",
      RACE == 7 ~ "Other Race",
      TRUE ~ "Multiple Races"
    )
  )
table(data2$Race)

unique(data2$HISPAN)
data2 <- data2 %>%
  mutate(
    Hispanic = case_when(
      HISPAN == 0 ~ "Not Hispanic",
      HISPAN %in% c(1, 2, 3, 4) ~ "Hispanic",
      TRUE ~ "Not Reported"
    )
  )
table(data2$Hispanic)

unique(data2$RACBLK)
data2 <- data2 %>%
  mutate(
    IsBlack = case_when(
      RACBLK == 1 ~ "Not Black",
      RACBLK == 2 ~ "Black",
      TRUE ~ "Missing/Blank"
    )
  )
table(data2$IsBlack)
```

# Rename education column and variables
N/A OR NO SCHOOLING
- 0: N/A or no schooling                                      - 1:	N/A
- 2:No schooling completed
GRADE SCHOOL & BELOW
- 10: Nursery school to grade 4                               - 11: Nursery school, preschool
- 12: Kindergarten                                            - 13: Grade 1, 2, 3, or 4
- 14: Grade 1                                                 - 15: Grade 2
- 16: Grade 3                                                 - 17: Grade 4
- 20: Grade 5, 6, 7, or 8                                     - 21: Grade 5 or 6
- 22: Grade 5                                                 - 23: Grade 6
- 24: Grade 7 or 8                                            - 25: Grade 7
- 26: Grade 8
HIGH SCHOOL OR GED
- 30: Grade 9                                                 - 40: Grade 10
- 50: Grade 11                                                - 60: Grade 12
- 61: 12th grade, no diploma                                  - 62: High school graduate or GED
- 63: Regular high school diploma                             - 64: GED or alternative credential
ASSOCIATE'S DEGREE OR SOME COLLEGE
- 65: Some college, but less than 1 year                      - 70: 1 year of college
- 71: 1 or more years of college credit, no degree.           - 80: 2 years of college          
- 81: Associate's degree, type not specified                  - 82: Associate's degree, occupational program
- 83: Associate's degree, academic program                    - 90: 3 years of college
- 100: 4 years of college                                     - 110: 5+ years of college
- 111: 6 years of college (6+ in 1960-1970)                   - 112: 7 years of college
- 113: 8+ years of college

- 101: Bachelor's degree
- 114: Master's degree
- 115: Professional degree beyond a bachelor's degree
- 116: Doctoral degree
- 999: Missing
```{r}
unique(data2$EDUCD)
data2 <- data2 %>%
  mutate(
    Education = case_when(
      EDUCD %in% c(0, 1, 2) ~ "N/A or No Schooling Completed",
      EDUCD %in% c(10, 11, 12, 13, 14, 15, 16, 17, 20, 21, 22, 23, 24, 25, 26) ~ "Grade School & Below",
      EDUCD %in% c(30, 40, 50, 60, 61, 62, 63, 64) ~ "High School or GED",
      EDUCD %in% c(65, 70, 71, 80, 81, 82, 83, 90, 100, 110, 111, 112, 113) ~ "Associate's Degree or Some College",
      EDUCD == 101 ~ "Bachelor's Degree",
      EDUCD == 114 ~ "Master's Degree",
      EDUCD == 115 ~ "Professional Degree Beyond a Bachelor's",
      EDUCD == 116 ~ "Doctoral Degree",
      TRUE ~ "Missing/Blank"
    )
  )
table(data2$Education)
```

# Rename degree column and variables
```{r}
unique(data2$DEGFIELD)
degreename <- c(
  "0"	= "N/A", "11"	= "Agriculture", "13" = "Environment and Natural Resources", "14" = "Architecture", 
  "15" = "Area, Ethnic, and Civilization Studies", "19" = "Communications", "20" = "Communication Technologies", 
  "21" = "Computer and Information Sciences", "22" = "Cosmetology Services and Culinary Arts", 
  "23" = "Education Administration and Teaching", "24" = "Engineering", "25" = "Engineering Technologies", 
  "26" = "Linguistics and Foreign Languages", "29" = "Family and Consumer Sciences", "32" = "Law", 
  "33" = "English Language, Literature, and Composition", "34" = "Liberal Arts and Humanities", 
  "35" = "Library Science", "36" = "Biology and Life Sciences", "37" = "Mathematics and Statistics", 
  "38" = "Military Technologies", "40" = "Interdisciplinary and Multi-Disciplinary Studies (General)", 
  "41" = "Physical Fitness, Parks, Recreation, and Leisure", "48" = "Philosophy and Religious Studies", 
  "49" = "Theology and Religious Vocations", "50" = "Physical Sciences", 
  "51" = "Nuclear, Industrial Radiology, and Biological Technologies", "52" = "Psychology", 
  "53" = "Criminal Justice and Fire Protection", "54" = "Public Affairs, Policy, and Social Work", 
  "55" = "Social Sciences", "56" = "Construction Services", "57" = "Electrical and Mechanic Repairs and Technologies",
  "58" = "Precision Production and Industrial Arts", "59" = "Transportation Sciences and Technologies", 
  "60" = "Fine Arts", "61" = "Medical and Health Sciences and Services", "62" = "Business", "64" = "History"
)
data2$Degree <- degreename[as.character(data2$DEGFIELD)]
table(data2$Degree)
```

# Create a new column for industry degree based on degree variables
```{r}
data2 <- data2 %>%
  mutate(
    DegreeIndustry = case_when(
      DEGFIELD %in% c(14, 19, 20, 21, 22, 24, 25, 56, 57, 58, 59, 60, 62) ~ "Corporate/Private Sector",
      DEGFIELD %in% c(13, 15, 23, 26, 33, 34, 35,  37, 40, 48, 49, 55, 64) ~ "Academia & Research",
      DEGFIELD %in% c(11, 29, 32, 38, 41, 53, 54) ~ "Legal & Public Sector",
      DEGFIELD %in% c(36, 50, 51, 52, 61) ~ "Healthcare",
      TRUE ~ "Missing/Blank"
    )
  )
table(data2$DegreeIndustry)
```

# Rename occupation column and variables
```{r}
unique(data2$OCC2010)
occupationname <- c(
  '10' = 'Chief executives and legislators/public administration', '20' = 'General and Operations Managers', 
  '30' = 'Managers in Marketing, Advertising, and Public Relations', '100' = 'Administrative Services Managers', 
  '110' = 'Computer and Information Systems Managers', '120' = 'Financial Managers', '130' = 'Human Resources Managers',
  '140' = 'Industrial Production Managers', '150' = 'Purchasing Managers', 
  '160' = 'Transportation, Storage, and Distribution Managers', '205' = 'Farmers, Ranchers, and Other Agricultural Managers', 
  '220' = 'Constructions Managers', '230' = 'Education Administrators', '300' = 'Architectural and Engineering Managers', 
  '310' = 'Food Service and Lodging Managers', '320' = 'Funeral Directors', '330' = 'Gaming Managers', 
  '350' = 'Medical and Health Services Managers', '360' = 'Natural Science Managers', 
  '410' = 'Property, Real Estate, and Community Association Managers', '420' = 'Social and Community Service Managers', 
  '430' = 'Managers, nec (including Postmasters)', '500' = 'Agents and Business Managers of Artists, Performers, and Athletes',
  '510' = 'Buyers and Purchasing Agents, Farm Products', '520' = 'Wholesale and Retail Buyers, Except Farm Products', 
  '530' = 'Purchasing Agents, Except Wholesale, Retail, and Farm Products', '540' = 'Claims Adjusters, Appraisers, Examiners, and Investigators', 
  '560' = 'Compliance Officers, Except Agriculture', '600' = 'Cost Estimators', '620' = 'Human Resources, Training, and Labor Relations Specialists',
  '700' = 'Logisticians', '710' = 'Management Analysts', '720' = 'Meeting and Convention Planners', 
  '730' = 'Other Business Operations and Management Specialists', '800' = 'Accountants and Auditors', '810' = 'Appraisers and Assessors of Real Estate',
  '820' = 'Budget Analysts', '830' = 'Credit Analysts', '840' = 'Financial Analysts', '850' = 'Personal Financial Advisors', 
  '860' = 'Insurance Underwriters', '900' = 'Financial Examiners', '910' = 'Credit Counselors and Loan Officers', 
  '930' = 'Tax Examiners and Collectors, and Revenue Agents', '940' = 'Tax Preparers', '950' = 'Financial Specialists, nec', 
  '1000' = 'Computer Scientists and Systems Analysts/Network systems Analysts/Web Developers', '1010' = 'Computer Programmers', 
  '1020' = 'Software Developers, Applications and Systems Software', '1050' = 'Computer Support Specialists', '1060' = 'Database Administrators', 
  '1100' = 'Network and Computer Systems Administrators', '1200' = 'Actuaries', '1220' = 'Operations Research Analysts', '1230' = 'Statisticians',
  '1240' = 'Mathematical science occupations, nec', '1300' = 'Architects, Except Naval', '1310' = 'Surveyors, Cartographers, and Photogrammetrists',
  '1320' = 'Aerospace Engineers', '1350' = 'Chemical Engineers', '1360' = 'Civil Engineers', '1400' = 'Computer Hardware Engineers', 
  '1410' = 'Electrical and Electronics Engineers', '1420' = 'Environmental Engineers', '1430' = 'Industrial Engineers, including Health and Safety',
  '1440' = 'Marine Engineers and Naval Architects', '1450' = 'Materials Engineers', '1460' = 'Mechanical Engineers', 
  '1520' = 'Petroleum, mining and geological engineers, including mining safety engineers', '1530' = 'Engineers, nec', '1540' = 'Drafters',
  '1550' = 'Engineering Technicians, Except Drafters', '1560' = 'Surveying and Mapping Technicians', '1600' = 'Agricultural and Food Scientists', 
  '1610' = 'Biological Scientists', '1640' = 'Conservation Scientists and Foresters', '1650' = 'Medical Scientists, and Life Scientists, All Other',
  '1700' = 'Astronomers and Physicists', '1710' = 'Atmospheric and Space Scientists', '1720' = 'Chemists and Materials Scientists', 
  '1740' = 'Environmental Scientists and Geoscientists', '1760' = 'Physical Scientists, nec', '1800' = 'Economists and market researchers', 
  '1820' = 'Psychologists', '1830' = 'Urban and Regional Planners', '1840' = 'Social Scientists, nec', 
  '1900' = 'Agricultural and Food Science Technicians', '1910' = 'Biological Technicians', '1920' = 'Chemical Technicians', 
  '1930' = 'Geological and Petroleum Technicians, and Nuclear Technicians', '1960' = 'Life, Physical, and Social Science Technicians, nec', 
  '1980' = 'Professional, Research, or Technical Workers, nec', '2000' = 'Counselors', '2010' = 'Social Workers', 
  '2020' = 'Community and Social Service Specialists, nec', '2040' = 'Clergy', '2050' = 'Directors, Religious Activities and Education', 
  '2060' = 'Religious Workers, nec', '2100' = 'Lawyers, and judges, magistrates, and other judicial workers', 
  '2140' = 'Paralegals and Legal Assistants', '2150' = 'Legal Support Workers, nec', '2200' = 'Postsecondary Teachers', 
  '2300' = 'Preschool and Kindergarten Teachers', '2310' = 'Elementary and Middle School Teachers', '2320' = 'Secondary School Teachers', 
  '2330' = 'Special Education Teachers', '2340' = 'Other Teachers and Instructors', '2400' = 'Archivists, Curators, and Museum Technicians', 
  '2430' = 'Librarians', '2440' = 'Library Technicians', '2540' = 'Teacher Assistants', '2550' = 'Education, Training, and Library Workers, nec', 
  '2600' = 'Artists and Related Workers', '2630' = 'Designers', '2700' = 'Actors, Producers, and Directors', 
  '2720' = 'Athletes, Coaches, Umpires, and Related Workers', '2740' = 'Dancers and Choreographers', '2750' = 'Musicians, Singers, and Related Workers',
  '2760' = 'Entertainers and Performers, Sports and Related Workers, All Other', '2800' = 'Announcers', 
  '2810' = 'Editors, News Analysts, Reporters, and Correspondents', '2825' = 'Public Relations Specialists', '2840' = 'Technical Writers', 
  '2850' = 'Writers and Authors', '2860' = 'Media and Communication Workers, nec', 
  '2900' = 'Broadcast and Sound Engineering Technicians and Radio Operators, and media and communication equipment workers, all other', 
  '2910' = 'Photographers', '2920' = 'Television, Video, and Motion Picture Camera Operators and Editors', '3000' = 'Chiropractors', 
  '3010' = 'Dentists', '3030' = 'Dieticians and Nutritionists', '3040' = 'Optometrists', '3050' = 'Pharmacists', '3060' = 'Physicians and Surgeons',
  '3110' = 'Physician Assistants', '3120' = 'Podiatrists', '3130' = 'Registered Nurses', '3140' = 'Audiologists', '3150' = 'Occupational Therapists',
  '3160' = 'Physical Therapists', '3200' = 'Radiation Therapists', '3210' = 'Recreational Therapists', '3220' = 'Respiratory Therapists', 
  '3230' = 'Speech Language Pathologists', '3240' = 'Therapists, nec', '3250' = 'Veterinarians', 
  '3260' = 'Health Diagnosing and Treating Practitioners, nec', '3300' = 'Clinical Laboratory Technologists and Technicians', 
  '3310' = 'Dental Hygienists', '3320' = 'Diagnostic Related Technologists and Technicians', '3400' = 'Emergency Medical Technicians and Paramedics',
  '3410' = 'Health Diagnosing and Treating Practitioner Support Technicians', '3500' = 'Licensed Practical and Licensed Vocational Nurses', 
  '3510' = 'Medical Records and Health Information Technicians', '3520' = 'Opticians, Dispensing', '3530' = 'Health Technologists and Technicians, nec',
  '3540' = 'Healthcare Practitioners and Technical Occupations, nec', '3600' = 'Nursing, Psychiatric, and Home Health Aides', 
  '3610' = 'Occupational Therapy Assistants and Aides', '3620' = 'Physical Therapist Assistants and Aides', '3630' = 'Massage Therapists', 
  '3640' = 'Dental Assistants', '3650' = 'Medical Assistants and Other Healthcare Support Occupations, nec', 
  '3700' = 'First-Line Supervisors of Correctional Officers', '3710' = 'First-Line Supervisors of Police and Detectives', 
  '3720' = 'First-Line Supervisors of Fire Fighting and Prevention Workers', '3730' = 'Supervisors, Protective Service Workers, All Other', 
  '3740' = 'Firefighters', '3750' = 'Fire Inspectors', '3800' = 'Sheriffs, Bailiffs, Correctional Officers, and Jailers', 
  '3820' = 'Police Officers and Detectives', '3900' = 'Animal Control', '3910' = 'Private Detectives and Investigators', 
  '3930' = 'Security Guards and Gaming Surveillance Officers', '3940' = 'Crossing Guards', '3950' = 'Law enforcement workers, nec', 
  '4000' = 'Chefs and Cooks', '4010' = 'First-Line Supervisors of Food Preparation and Serving Workers', '4030' = 'Food Preparation Workers', 
  '4040' = 'Bartenders', '4050' = 'Combined Food Preparation and Serving Workers, Including Fast Food', 
  '4060' = 'Counter Attendant, Cafeteria, Food Concession, and Coffee Shop', '4110' = 'Waiters and Waitresses', '4120' = 'Food Servers, Nonrestaurant',
  '4130' = 'Food preparation and serving related workers, nec', '4140' = 'Dishwashers', 
  '4150' = 'Host and Hostesses, Restaurant, Lounge, and Coffee Shop', '4200' = 'First-Line Supervisors of Housekeeping and Janitorial Workers', 
  '4210' = 'First-Line Supervisors of Landscaping, Lawn Service, and Groundskeeping Workers', '4220' = 'Janitors and Building Cleaners', 
  '4230' = 'Maids and Housekeeping Cleaners', '4240' = 'Pest Control Workers', '4250' = 'Grounds Maintenance Workers', 
  '4300' = 'First-Line Supervisors of Gaming Workers', '4320' = 'First-Line Supervisors of Personal Service Workers', '4340' = 'Animal Trainers', 
  '4350' = 'Nonfarm Animal Caretakers', '4400' = 'Gaming Services Workers', '4420' = 'Ushers, Lobby Attendants, and Ticket Takers', 
  '4430' = 'Entertainment Attendants and Related Workers, nec', '4460' = 'Funeral Service Workers and Embalmers', '4500' = 'Barbers', 
  '4510' = 'Hairdressers, Hairstylists, and Cosmetologists', '4520' = 'Personal Appearance Workers, nec', 
  '4530' = 'Baggage Porters, Bellhops, and Concierges', '4540' = 'Tour and Travel Guides', '4600' = 'Childcare Workers', '4610' = 'Personal Care Aides',
  '4620' = 'Recreation and Fitness Workers', '4640' = 'Residential Advisors', '4650' = 'Personal Care and Service Workers, All Other', 
  '4700' = 'First-Line Supervisors of Sales Workers', '4720' = 'Cashiers', '4740' = 'Counter and Rental Clerks', '4750' = 'Parts Salespersons', 
  '4760' = 'Retail Salespersons', '4800' = 'Advertising Sales Agents', '4810' = 'Insurance Sales Agents', 
  '4820' = 'Securities, Commodities, and Financial Services Sales Agents', '4830' = 'Travel Agents', 
  '4840' = 'Sales Representatives, Services, All Other', '4850' = 'Sales Representatives, Wholesale and Manufacturing', 
  '4900' = 'Models, Demonstrators, and Product Promoters', '4920' = 'Real Estate Brokers and Sales Agents', '4930' = 'Sales Engineers', 
  '4940' = 'Telemarketers', '4950' = 'Door-to-Door Sales Workers, News and Street Vendors, and Related Workers', 
  '4965' = 'Sales and Related Workers, All Other', '5000' = 'First-Line Supervisors of Office and Administrative Support Workers', 
  '5010' = 'Switchboard Operators, Including Answering Service', '5020' = 'Telephone Operators', 
  '5030' = 'Communications Equipment Operators, All Other', '5100' = 'Bill and Account Collectors', '5110' = 'Billing and Posting Clerks', 
  '5120' = 'Bookkeeping, Accounting, and Auditing Clerks', '5130' = 'Gaming Cage Workers', '5140' = 'Payroll and Timekeeping Clerks', 
  '5150' = 'Procurement Clerks', '5160' = 'Bank Tellers', '5165' = 'Financial Clerks, nec', '5200' = 'Brokerage Clerks', 
  '5220' = 'Court, Municipal, and License Clerks', '5230' = 'Credit Authorizers, Checkers, and Clerks', '5240' = 'Customer Service Representatives',
  '5250' = 'Eligibility Interviewers, Government Programs', '5260' = 'File Clerks', '5300' = 'Hotel, Motel, and Resort Desk Clerks', 
  '5310' = 'Interviewers, Except Eligibility and Loan', '5320' = 'Library Assistants, Clerical', '5330' = 'Loan Interviewers and Clerks', 
  '5340' = 'New Account Clerks', '5350' = 'Correspondent clerks and order clerks', 
  '5360' = 'Human Resources Assistants, Except Payroll and Timekeeping', '5400' = 'Receptionists and Information Clerks', 
  '5410' = 'Reservation and Transportation Ticket Agents and Travel Clerks', '5420' = 'Information and Record Clerks, All Other', 
  '5500' = 'Cargo and Freight Agents', '5510' = 'Couriers and Messengers', '5520' = 'Dispatchers', '5530' = 'Meter Readers, Utilities', 
  '5540' = 'Postal Service Clerks', '5550' = 'Postal Service Mail Carriers', 
  '5560' = 'Postal Service Mail Sorters, Processors, and Processing Machine Operators', '5600' = 'Production, Planning, and Expediting Clerks', 
  '5610' = 'Shipping, Receiving, and Traffic Clerks', '5620' = 'Stock Clerks and Order Fillers', 
  '5630' = 'Weighers, Measurers, Checkers, and Samplers, Recordkeeping', '5700' = 'Secretaries and Administrative Assistants', 
  '5800' = 'Computer Operators', '5810' = 'Data Entry Keyers', '5820' = 'Word Processors and Typists', 
  '5840' = 'Insurance Claims and Policy Processing Clerks', '5850' = 'Mail Clerks and Mail Machine Operators, Except Postal Service', 
  '5860' = 'Office Clerks, General', '5900' = 'Office Machine Operators, Except Computer', '5910' = 'Proofreaders and Copy Markers', 
  '5920' = 'Statistical Assistants', '5940' = 'Office and administrative support workers, nec', 
  '6005' = 'First-Line Supervisors of Farming, Fishing, and Forestry Workers', '6010' = 'Agricultural Inspectors', 
  '6040' = 'Graders and Sorters, Agricultural Products', '6050' = 'Agricultural workers, nec', '6100' = 'Fishing and hunting workers', 
  '6120' = 'Forest and Conservation Workers', '6130' = 'Logging Workers', 
  '6200' = 'First-Line Supervisors of Construction Trades and Extraction Workers', '6210' = 'Boilermakers', 
  '6220' = 'Brickmasons, Blockmasons, and Stonemasons', '6230' = 'Carpenters', '6240' = 'Carpet, Floor, and Tile Installers and Finishers', 
  '6250' = 'Cement Masons, Concrete Finishers, and Terrazzo Workers', '6260' = 'Construction Laborers', 
  '6300' = 'Paving, Surfacing, and Tamping Equipment Operators', 
  '6320' = 'Construction equipment operators except paving, surfacing, and tamping equipment operators', 
  '6330' = 'Drywall Installers, Ceiling Tile Installers, and Tapers', '6355' = 'Electricians', '6360' = 'Glaziers', '6400' = 'Insulation Workers',
  '6420' = 'Painters, Construction and Maintenance', '6430' = 'Paperhangers', '6440' = 'Pipelayers, Plumbers, Pipefitters, and Steamfitters', 
  '6460' = 'Plasterers and Stucco Masons', '6500' = 'Reinforcing Iron and Rebar Workers', '6515' = 'Roofers', 
  '6520' = 'Sheet Metal Workers, metal-working', '6530' = 'Structural Iron and Steel Workers', '6600' = 'Helpers, Construction Trades', 
  '6660' = 'Construction and Building Inspectors', '6700' = 'Elevator Installers and Repairers', '6710' = 'Fence Erectors', 
  '6720' = 'Hazardous Materials Removal Workers', '6730' = 'Highway Maintenance Workers', 
  '6740' = 'Rail-Track Laying and Maintenance Equipment Operators', '6765' = 'Construction workers, nec', 
  '6800' = 'Derrick, rotary drill, and service unit operators, and roustabouts, oil, gas, and mining', '6820' = 'Earth Drillers, Except Oil and Gas',
  '6830' = 'Explosives Workers, Ordnance Handling Experts, and Blasters', '6840' = 'Mining Machine Operators', '6940' = 'Extraction workers, nec',
  '7000' = 'First-Line Supervisors of Mechanics, Installers, and Repairers', '7010' = 'Computer, Automated Teller, and Office Machine Repairers', 
  '7020' = 'Radio and Telecommunications Equipment Installers and Repairers', '7030' = 'Avionics Technicians', 
  '7040' = 'Electric Motor, Power Tool, and Related Repairers', 
  '7100' = 'Electrical and electronics repairers, transportation equipment, and industrial and utility', 
  '7110' = 'Electronic Equipment Installers and Repairers, Motor Vehicles', 
  '7120' = 'Electronic Home Entertainment Equipment Installers and Repairers', '7125' = 'Electronic Repairs, nec', 
  '7130' = 'Security and Fire Alarm Systems Installers', '7140' = 'Aircraft Mechanics and Service Technicians', 
  '7150' = 'Automotive Body and Related Repairers', '7160' = 'Automotive Glass Installers and Repairers', 
  '7200' = 'Automotive Service Technicians and Mechanics', '7210' = 'Bus and Truck Mechanics and Diesel Engine Specialists', 
  '7220' = 'Heavy Vehicle and Mobile Equipment Service Technicians and Mechanics', '7240' = 'Small Engine Mechanics', 
  '7260' = 'Vehicle and Mobile Equipment Mechanics, Installers, and Repairers, nec', '7300' = 'Control and Valve Installers and Repairers', 
  '7315' = 'Heating, Air Conditioning, and Refrigeration Mechanics and Installers', '7320' = 'Home Appliance Repairers', 
  '7330' = 'Industrial and Refractory Machinery Mechanics', '7340' = 'Maintenance and Repair Workers, General', 
  '7350' = 'Maintenance Workers, Machinery', '7360' = 'Millwrights', '7410' = 'Electrical Power-Line Installers and Repairers', 
  '7420' = 'Telecommunications Line Installers and Repairers', '7430' = 'Precision Instrument and Equipment Repairers', 
  '7510' = 'Coin, Vending, and Amusement Machine Servicers and Repairers', '7540' = 'Locksmiths and Safe Repairers', 
  '7550' = 'Manufactured Building and Mobile Home Installers', '7560' = 'Riggers', '7610' = 'Helpers--Installation, Maintenance, and Repair Workers',
  '7630' = 'Other Installation, Maintenance, and Repair Workers Including Wind Turbine Service Technicians, and Commercial Divers, and Signal and Track Switch Repairers', '7700' = 'First-Line Supervisors of Production and Operating Workers', 
  '7710' = 'Aircraft Structure, Surfaces, Rigging, and Systems Assemblers', '7720' = 'Electrical, Electronics, and Electromechanical Assemblers', 
  '7730' = 'Engine and Other Machine Assemblers', '7740' = 'Structural Metal Fabricators and Fitters', '7750' = 'Assemblers and Fabricators, nec',
  '7800' = 'Bakers', '7810' = 'Butchers and Other Meat, Poultry, and Fish Processing Workers', 
  '7830' = 'Food and Tobacco Roasting, Baking, and Drying Machine Operators and Tenders', 
  '7840' = 'Food Batchmakers', '7850' = 'Food Cooking Machine Operators and Tenders', '7855' = 'Food Processing, nec', 
  '7900' = 'Computer Control Programmers and Operators', 
  '7920' = 'Extruding and Drawing Machine Setters, Operators, and Tenders, Metal and Plastic', 
  '7930' = 'Forging Machine Setters, Operators, and Tenders, Metal and Plastic', 
  '7940' = 'Rolling Machine Setters, Operators, and Tenders, metal and Plastic', 
  '7950' = 'Cutting, Punching, and Press Machine Setters, Operators, and Tenders, Metal and Plastic', 
  '7960' = 'Drilling and Boring Machine Tool Setters, Operators, and Tenders, Metal and Plastic', 
  '8000' = 'Grinding, Lapping, Polishing, and Buffing Machine Tool Setters, Operators, and Tenders, Metal and Plastic', 
  '8010' = 'Lathe and Turning Machine Tool Setters, Operators, and Tenders, Metal and Plastic', '8030' = 'Machinists', 
  '8040' = 'Metal Furnace Operators, Tenders, Pourers, and Casters', '8060' = 'Model Makers and Patternmakers, Metal and Plastic', 
  '8100' = 'Molders and Molding Machine Setters, Operators, and Tenders, Metal and Plastic', '8130' = 'Tool and Die Makers', 
  '8140' = 'Welding, Soldering, and Brazing Workers', '8150' = 'Heat Treating Equipment Setters, Operators, and Tenders, Metal and Plastic', 
  '8200' = 'Plating and Coating Machine Setters, Operators, and Tenders, Metal and Plastic', '8210' = 'Tool Grinders, Filers, and Sharpeners', 
  '8220' = 'Metal workers and plastic workers, nec', '8230' = 'Bookbinders, Printing Machine Operators, and Job Printers', 
  '8250' = 'Prepress Technicians and Workers', '8300' = 'Laundry and Dry-Cleaning Workers', 
  '8310' = 'Pressers, Textile, Garment, and Related Materials', '8320' = 'Sewing Machine Operators', '8330' = 'Shoe and Leather Workers and Repairers',
  '8340' = 'Shoe Machine Operators and Tenders', '8350' = 'Tailors, Dressmakers, and Sewers', 
  '8400' = 'Textile bleaching and dyeing, and cutting machine setters, operators, and tenders', 
  '8410' = 'Textile Knitting and Weaving Machine Setters, Operators, and Tenders', 
  '8420' = 'Textile Winding, Twisting, and Drawing Out Machine Setters, Operators, and Tenders', '8450' = 'Upholsterers', 
  '8460' = 'Textile, Apparel, and Furnishings workers, nec', '8500' = 'Cabinetmakers and Bench Carpenters', '8510' = 'Furniture Finishers', 
  '8530' = 'Sawing Machine Setters, Operators, and Tenders, Wood', '8540' = 'Woodworking Machine Setters, Operators, and Tenders, Except Sawing', 
  '8550' = 'Woodworkers including model makers and patternmakers, nec', '8600' = 'Power Plant Operators, Distributors, and Dispatchers', 
  '8610' = 'Stationary Engineers and Boiler Operators', '8620' = 'Water Wastewater Treatment Plant and System Operators', 
  '8630' = 'Plant and System Operators, nec', '8640' = 'Chemical Processing Machine Setters, Operators, and Tenders', 
  '8650' = 'Crushing, Grinding, Polishing, Mixing, and Blending Workers', '8710' = 'Cutting Workers', 
  '8720' = 'Extruding, Forming, Pressing, and Compacting Machine Setters, Operators, and Tenders', 
  '8730' = 'Furnace, Kiln, Oven, Drier, and Kettle Operators and Tenders', '8740' = 'Inspectors, Testers, Sorters, Samplers, and Weighers', 
  '8750' = 'Jewelers and Precious Stone and Metal Workers', '8760' = 'Medical, Dental, and Ophthalmic Laboratory Technicians', 
  '8800' = 'Packaging and Filling Machine Operators and Tenders', '8810' = 'Painting Workers and Dyers', 
  '8830' = 'Photographic Process Workers and Processing Machine Operators', '8850' = 'Adhesive Bonding Machine Operators and Tenders', 
  '8860' = 'Cleaning, Washing, and Metal Pickling Equipment Operators and Tenders', '8910' = 'Etchers, Engravers, and Lithographers', 
  '8920' = 'Molders, Shapers, and Casters, Except Metal and Plastic', '8930' = 'Paper Goods Machine Setters, Operators, and Tenders', 
  '8940' = 'Tire Builders', '8950' = 'Helpers--Production Workers', 
  '8965' = 'Other production workers including semiconductor processors and cooling and freezing equipment operators', 
  '9000' = 'Supervisors of Transportation and Material Moving Workers', '9030' = 'Aircraft Pilots and Flight Engineers', 
  '9040' = 'Air Traffic Controllers and Airfield Operations Specialists', '9050' = 'Flight Attendants and Transportation Workers and Attendants', 
  '9100' = 'Bus and Ambulance Drivers and Attendants', '9130' = 'Driver/Sales Workers and Truck Drivers', '9140' = 'Taxi Drivers and Chauffeurs', 
  '9150' = 'Motor Vehicle Operators, All Other', '9200' = 'Locomotive Engineers and Operators', '9230' = 'Railroad Brake, Signal, and Switch Operators',
  '9240' = 'Railroad Conductors and Yardmasters', '9260' = 'Subway, Streetcar, and Other Rail Transportation Workers', 
  '9300' = 'Sailors and marine oilers, and ship engineers', '9310' = 'Ship and Boat Captains and Operators', '9350' = 'Parking Lot Attendants', 
  '9360' = 'Automotive and Watercraft Service Attendants', '9410' = 'Transportation Inspectors', '9420' = 'Transportation workers, nec', 
  '9510' = 'Crane and Tower Operators', '9520' = 'Dredge, Excavating, and Loading Machine Operators', 
  '9560' = 'Conveyor operators and tenders, and hoist and winch operators', '9600' = 'Industrial Truck and Tractor Operators', 
  '9610' = 'Cleaners of Vehicles and Equipment', '9620' = 'Laborers and Freight, Stock, and Material Movers, Hand', 
  '9630' = 'Machine Feeders and Offbearers', '9640' = 'Packers and Packagers, Hand', '9650' = 'Pumping Station Operators', 
  '9720' = 'Refuse and Recyclable Material Collectors', '9750' = 'Material moving workers, nec', 
  '9800' = 'Military Officer Special and Tactical Operations Leaders', '9810' = 'First-Line Enlisted Military Supervisors', 
  '9820' = 'Military Enlisted Tactical Operations and Air/Weapons Specialists and Crew Members', '9830' = 'Military, Rank Not Specified', 
  '9920' = 'Unemployed, with No Work Experience in the Last 5 Years or Earlier or Never Worked'
)
data2$Occupation <- occupationname[as.character(data2$OCC2010)]
table(data2$Occupation)
```

# Create new columns for industry work and leadership based on occupation variables
Occupation Code Breakdown - https://docs.google.com/spreadsheets/d/1z71zMLqKpXTiBQCN-e6C1wbjrAaZ-r8q/edit?usp=share_link&ouid=116402902036445997313&rtpof=true&sd=true
```{r}
data2 <- data2 %>%
  mutate(
    WorkIndustry = case_when(
      OCC2010 %in% c(10, 20, 30, 100, 110, 120, 130, 140, 150, 160, 220, 300, 310, 330, 410, 500, 510, 520, 530, 540, 600, 620, 700, 710, 720, 730, 800,
                     810, 820, 830, 840, 850, 860, 900, 910, 940, 950, 1000, 1010, 1020, 1050, 1060, 1100, 1200, 1300, 1310, 1320, 1350, 1360, 1400,
                     1410, 1430, 1440, 1450, 1460, 1520, 1530, 1540, 1550, 1560, 1800, 1830, 1930, 2600, 2630, 2700, 2720, 2740, 2750, 2760, 2800, 2810,
                     2825, 2840, 2850, 2860, 2900, 2910, 2920, 4000, 4010, 4030, 4040, 4050, 4060, 4110, 4120, 4130, 4140, 4150, 4200, 4210, 4220, 4230,
                     4240, 4250, 4300, 4320, 4340, 4350, 4400, 4420, 4430, 4460, 4500, 4510, 4520, 4530, 4540, 4600, 4610, 4620, 4640, 4650, 4700, 4720,
                     4740, 4750, 4760, 4800, 4810, 4820, 4830, 4840, 4850, 4900, 4920, 4930, 4940, 4950, 4965, 5000, 5010, 5020, 5030, 5100, 5110, 5120,
                     5130, 5140, 5150, 5160, 5165, 5200, 5230, 5240, 5260, 5300, 5310, 5320, 5330, 5340, 5350, 5360, 5400, 5410, 5420, 5500, 5510, 5520,
                     5530, 5600, 5610, 5620, 5630, 5700, 5800, 5810, 5820, 5840, 5850, 5860, 5900, 5910, 5940, 6005, 6040, 6050, 6100, 6130, 6200, 6210,
                     6220, 6230, 6240, 6250, 6260, 6300, 6320, 6330, 6355, 6360, 6400, 6420, 6430, 6440, 6460, 6500, 6515, 6520, 6530, 6600, 6660, 6700,
                     6710, 6720, 6765, 6800, 6820, 6830, 6840, 6940, 7000, 7010, 7020, 7030, 7040, 7100, 7110, 7120, 7125, 7130, 7140, 7150, 7160, 7200,
                     7210, 7220, 7240, 7260, 7300, 7315, 7320, 7330, 7340, 7350, 7360, 7410, 7420, 7430, 7510, 7540, 7550, 7560, 7610, 7630, 7700, 7710,
                     7720, 7730, 7740, 7750, 7800, 7810, 7830, 7840, 7850, 7855, 7900, 7920, 7930, 7940, 7950, 7960, 8000, 8010, 8030, 8040, 8060, 8100,
                     8130, 8140, 8150, 8200, 8210, 8220, 8230, 8250, 8300, 8310, 8320, 8330, 8340, 8350, 8400, 8410, 8420, 8450, 8460, 8500, 8510, 8530,
                     8540, 8550, 8600, 8610, 8630, 8640, 8650, 8710, 8720, 8730, 8740, 8750, 8800, 8810, 8830, 8850, 8860, 8910, 8920, 8930, 8940, 8950,
                     8965, 9000, 9030, 9050, 9130, 9140, 9150, 9200, 9230, 9240, 9300, 9310, 9350, 9360, 9420, 9510, 9520, 9560, 9600, 9610, 9620, 9630,
                     9640, 9650, 9750) ~ "Corporate/Private Sector",
      OCC2010 %in% c(230, 360, 1220, 1230, 1240, 1420, 1600, 1610, 1640, 1700, 1710, 1720, 1740, 1760, 1820, 1840, 1900, 1910, 1920, 1960, 1980, 2200,
                     2300, 2310, 2320, 2330, 2340, 2400, 2430, 2440, 2540, 2550, 5920) ~ "Academia & Research",
      OCC2010 %in% c(205, 420, 430, 560, 930, 2000, 2010, 2020, 2040, 2050, 2060, 2100, 2140, 2150, 3700, 3710, 3720, 3730, 3740, 3750, 3800, 3820,
                     3900, 3910, 3930, 3940, 3950, 5220, 5250, 5540, 5550, 5560, 6010, 6120, 6730, 6740, 8620, 9040, 9100, 9260, 9410, 9720, 9800, 9810,
                     9820, 9830) ~ "Legal & Public Sector",
      OCC2010 %in% c(320, 350, 1650, 3000, 3010, 3030, 3040, 3050, 3060, 3110, 3120, 3130, 3140, 3150, 3160, 3200, 3210, 3220, 3230, 3240, 3250, 3260,
                     3300, 3310, 3320, 3400, 3410, 3500, 3510, 3520, 3530, 3540, 3600, 3610, 3620, 3630, 3640, 3650, 8760) ~ "Healthcare",
      TRUE ~ "Unemployed"
    )
  )
table(data2$WorkIndustry)

data2 <- data2 %>%
  mutate(
    Leadership = case_when(
      OCC2010 %in% c(10, 20, 30, 100, 110, 120, 130, 140, 150, 160, 220, 230, 300, 350, 360, 420, 430, 1220, 1230, 1240, 1300, 1320, 1350, 1360, 1400,
                     1410, 1420, 1430, 1440, 1450, 1600, 1610, 1640, 1650, 1700, 1710, 1720, 1740, 1760, 1800, 1820, 1830, 1840, 2000, 2010, 2040, 2050,
                     2100, 2200, 2400, 2430, 3000, 3010, 3030, 3040, 3050, 3060, 3110, 3120, 3140, 3150, 3160, 3200, 3210, 3220, 3230, 3240, 3250, 3260,
                     3540, 4820, 4930, 9030, 9040, 9800) ~ "Yes",
      TRUE ~ "No"
    )
  )
table(data2$Leadership)
```

# Adjust income data for inflation (based on U.S. Bureau of Labor Statistics)
```{r}
unique(data2$INCTOT)
unique(data2$INCWAGE)

cpi <- data.frame(
  year = 2010:2023,
  cpi = c(218.056, 224.939, 229.594, 232.957, 236.736, 237.017, 240.007, 245.120, 251.107, 255.657, 258.811, 270.970, 292.655, 305.139)
)

data2 <- merge(data2, cpi, by.x = "YEAR", by.y = "year")
data2$cpi23 <- 305.139

data2$totalincome <- data2$INCTOT * data2$ADJUST * (data2$cpi23 / data2$cpi)
data2$wageincome <- data2$INCWAGE * data2$ADJUST * (data2$cpi23 / data2$cpi)

table(data2$totalincome)
table(data2$wageincome)
```

# Take a quick look at the data again
```{r}
head(data2, 10)
glimpse(data2)
```

# Now subset the data we will be working with 
```{r}
cleandata <- data2 %>%
  select(
    "Year" = YEAR, "Region Code" = REGION, "Region" = Region, "State Code" = STATEFIP, "State" = State,
    "Sex Code" = SEX, "Sex" = Sex, "Age" = AGE,
    "Race Code" = RACE, "Race" = Race, "Hispanic Code" = HISPAN, "Hispanic" = Hispanic, "Black Code" = RACBLK, "Black" = IsBlack,
    "Educational Code" = EDUCD, "Educational Attainment" = Education,
    "Degree Code" = DEGFIELD, "Degree" = Degree, "Degree Industry" = DegreeIndustry,
    "Occupation Code" = OCC2010, "Occupation" = Occupation, "Work Industry" = WorkIndustry,
    "Leadership" = Leadership, "Unadjusted Total Income" = INCTOT, "Total Income" = totalincome, "Unadjusted Wage Income" = INCWAGE, "Wage Income" = wageincome
  )
head(cleandata, 10)
glimpse(cleandata)
```

# Finally print the cleaned data
```{r}
write.csv(cleandata, "Clean 2010 - 2023 ACS IPUMS Data.csv", row.names = FALSE)
```





